---
title: "RMS Titanic"
author: "Angela Sharer"
date: 2020-07-11
output:
  github_document:
    toc: true
---

```{r include = FALSE}
library(tidyverse)

df_titanic <- as_tibble(Titanic)
```

*Background*: The RMS Titanic sank on its maiden voyage in 1912; about 67% of its passengers died.

# First Look
<!-- -------------------------------------------------- -->

__q1__ Perform a glimpse of `df_titanic`. What variables are in this dataset?

```{r q1-task - variables}
## TASK: Perform a `glimpse` of df_titanic
glimpse(df_titanic)
```

**Observations**:

- Class (this includes ticketed passengers' cabin class, e.g. "1st," and also includes employees, e.g. "Crew")
- Sex ("Male," "Female")
- Age (this is not numerical--it's a string, e.g. "Child")
- Survived (again, this is a string, e.g. "No")
- n (this is the only numerical data in df_titanic, and looks to be a count of people of this class & sex & age & survival status)

__q2__ Skim the [Wikipedia article](https://en.wikipedia.org/wiki/RMS_Titanic) on the RMS Titanic, and look for a total count of passengers. Compare against the total computed below. Are there any differences? Are those differences large or small? What might account for those differences?

```{r q2-summary - people count}
## NOTE: No need to edit! We'll cover how to
## do this calculation in a later exercise.
df_titanic %>% summarize(total = sum(n))
```

**Observations**:

- According to Wikipedia, the Titanic carried 2,224 passengers and crew (estimated), and our sum is close to this--2,201, or only 23 people short of Wikipedia's estimate.
- I would characterize this difference as small. 23 people is only about 1% of 2,224, so if Wikipedia's larger figure is correct, we are only missing 1% of passengers and crew.
- As for the differences, the number of deaths are disputed, due to some passengers canceling their trip at the last minute (or perhaps not making the sailing), but still being included in the ship's passenger manifest; also, according to Wikipedia, some passengers traveled under aliases, and were counted twice in the casualties list.

__q3__ Create a plot showing the count of passengers who *did* survive, along with aesthetics for `Class` and `Sex`. Document your observations below.

*Note*: There are many ways to do this.

```{r q3-task - class and sex counts, focused on class}
## TASK: Visualize counts against `Class` and `Sex`
ggplot(df_titanic) +
  geom_col(mapping = aes(x = Class, y = n, fill = Sex))
```

**Observations**:

- The Titanic had more 3rd class passengers than 1st and 2nd class passengers combined.
- 2nd class is the smallest of the classes. I would have expected 2nd to be larger than 1st, but perhaps, as a luxury liner, Titanic was optimizing for a relatively large 1st class, compared to 2nd? 
- There were more crew members than any single passenger class (but as a whole, there were more passengers than crew).
- The vast majority of the crew were male.
- Most passengers were male.

```{r q3-task - class and sex counts, focused on class proportion}
## TASK: Visualize counts against `Class` and `Sex`
ggplot(df_titanic) +
  geom_col(mapping = aes(x = Class, y = n, fill = Sex), position = "fill")
```

**Observations**:

- 1st class comes the closest of any other class group to an even balance of males and females, but it too had more males than females.
- I'm wondering if 1st class passengers were more likely to be couples or families for some reason, and 2nd and 3rd class had more single men--what social factors would have caused this?


```{r q3-task - class and sex counts, focused on sex}
## TASK: Visualize counts against `Class` and `Sex`

ggplot(df_titanic) +
  geom_col(mapping = aes(x = Sex, y = n, fill = Class))
```

**Observations**:

- The ship's male crew members alone outnumber the women and girls.

```{r q3-task - class and sex counts, focused on sex, by proportion}
## TASK: Visualize counts against `Class` and `Sex`

ggplot(df_titanic) +
  geom_col(mapping = aes(x = Sex, y = n, fill = Class), position = "fill")
```

**Observations**:

- About half of the men/boys on the Titanic were crew members.


# Deeper Look
<!-- -------------------------------------------------- -->

Raw counts give us a sense of totals, but they are not as useful for understanding differences between groups. This is because the differences we see in counts could be due to either the relative size of the group OR differences in outcomes for those groups. To make comparisons between groups, we should also consider *proportions*.[1]

The following code computes proportions within each `Class, Sex, Age` group.

```{r compute-prop}
## NOTE: No need to edit! We'll cover how to
## do this calculation in a later exercise.
df_prop <-
  df_titanic %>%
  group_by(Class, Sex, Age) %>%
  mutate(
    Total = sum(n),
    Prop = n / Total
  ) %>%
  ungroup()
df_prop
```

__q4__ Replicate your visual from q3, but display `Prop` in place of `n`. Document your observations, and note any new/different observations you make in comparison with q3.

```{r q4-task - class and sex proportions}
```

**Observations**:

- Write your observations here.

__q5__ Create a plot showing the group-proportion of passengers who *did* survive, along with aesthetics for `Class`, `Sex`, *and* `Age`. Document your observations below.

*Hint*: Don't forget that you can use `facet_grid` to help consider additional variables!

```{r q5-task survivors by class, sex, and age}
```

**Observations**:

- Write your observations here.


## Purpose and Reading
*Purpose*: Most datasets have at least a few variables. Part of our task in analyzing a dataset is to understand trends as they vary across these different variables. Unless we're careful and thorough, we can easily miss these patterns. In this challenge you'll analyze a dataset with a small number of categorical variables and try to find differences among the groups.

*Reading*: (Optional) [Wikipedia article](https://en.wikipedia.org/wiki/RMS_Titanic) on the RMS Titanic.

<!-- include-rubric -->
# Grading Rubric
<!-- -------------------------------------------------- -->

Unlike exercises, **challenges will be graded**. The following rubrics define how you will be graded, both on an individual and team basis.

## Individual
<!-- ------------------------- -->

| Category | Unsatisfactory | Satisfactory |
|----------|----------------|--------------|
| Effort | Some task __q__'s left unattempted | All task __q__'s attempted |
| Observed | Did not document observations | Documented observations based on analysis |
| Supported | Some observations not supported by analysis | All observations supported by analysis (table, graph, etc.) |
| Code Styled | Violations of the [style guide](https://style.tidyverse.org/) hinder readability | Code sufficiently close to the [style guide](https://style.tidyverse.org/) |

## Team
<!-- ------------------------- -->

| Category | Unsatisfactory | Satisfactory |
|----------|----------------|--------------|
| Documented | No team contributions to Wiki | Team contributed to Wiki |
| Referenced | No team references in Wiki | At least one reference in Wiki to member report(s) |
| Relevant | References unrelated to assertion, or difficult to find related analysis based on reference text | Reference text clearly points to relevant analysis |

## Due Date
<!-- ------------------------- -->

All the deliverables stated in the rubrics above are due on the day of the class discussion of that exercise. See the [Syllabus](https://docs.google.com/document/d/1jJTh2DH8nVJd2eyMMoyNGroReo0BKcJrz1eONi3rPSc/edit?usp=sharing) for more information.

# Notes
<!-- -------------------------------------------------- -->

[1] This is basically the same idea as [Dimensional Analysis](https://en.wikipedia.org/wiki/Dimensional_analysis); computing proportions is akin to non-dimensionalizing a quantity.
