---
title: "Gapminder"
author: "Angela Sharer"
date: 2020-07-24
output:
  github_document:
    toc: true
---

*Purpose*: Learning to do EDA well takes practice! In this challenge you'll further practice EDA by first completing a guided exploration, then by conducting your own investigation. This challenge will also give you a chance to use the wide variety of visual tools we've been learning.


```{r setup}
library(tidyverse)
library(gapminder)

```

*Background*: [Gapminder](https://www.gapminder.org/about-gapminder/) is an independent organization that seeks to education people about the state of the world. They promote a "fact-based worldview" by focusing on data. The dataset we'll study in this challenge is from Gapminder.

# Guided EDA
<!-- -------------------------------------------------- -->

First, we'll go through a round of *guided EDA*. Try to pay attention to the high-level process we're going through---after this guided round you'll be responsible for doing another cycle of EDA on your own!

__q0__ Perform your "first checks" on the dataset. What variables are in this
dataset?

```{r q0-task}
## TASK: Do your "first checks" here!
glimpse(gapminder)
```

**Observations**:

- `country`: Country (factor)
- `continent`: Continent (factor)
- `year`: Year (integer; 4 digits)
- `lifeExp`: Life Expectancy in years (double)
- `pop`: Population (int)
- `gdpPercap`: Gross Domestic Product per Capita

__q1__ Determine the most and least recent years in the `gapminder` dataset.

```{r q1-task}
## TASK: Find the largest and smallest values of `year` in `gapminder`

year_max <- max(select(gapminder, year))
year_min <- min(select(gapminder, year))

```

Use the following test to check your work.

```{r q1-tests}
## NOTE: No need to change this
assertthat::assert_that(year_max %% 7 == 5)
assertthat::assert_that(year_max %% 3 == 0)
assertthat::assert_that(year_min %% 7 == 6)
assertthat::assert_that(year_min %% 3 == 2)
print("Nice!")
```

__q2__ Filter on years matching `year_min`, and make a plot of the GDE per capita against continent. Choose an appropriate `geom_` to visualize the data. What observations can you make?

You may encounter difficulties in visualizing these data; if so document your challenges and attempt to produce the most informative visual you can.

```{r q2-task}
## TASK: Create a visual of gdpPercap vs continent
gapminder %>%
  filter(year == year_min) %>%
  ggplot(aes(gdpPercap, continent)) +
    geom_boxplot() +
    coord_trans(x = "log") +
    labs(title = "GDP per capita by continent", x = "GDP per capita", y = "Continent")

```

**Observations**:

- In 1952, Oceania had the highest mean GPD per capita, followed by Europe, the Americas, Asia, and lastly Africa.
- Asia has the greatest spread of GDP per capita of any continent, having one country with a GDP that is much higher than any other country this year.

**Difficulties & Approaches**:

- Difficult to differentiate between GDP values near zero, where many are clustered. Used a boxplot for a high-level summary and flipped the axes to get greater resolution in GDP.  
- This still wasn't enough. Scaled the x-axis to be logarithmic.

__q3__ You should have found at least three outliers in q2. Identify those outliers (figure out which countries they are).

```{r q3-task}
## TASK: Identify the outliers from q2
outliers_1952 <-
  gapminder %>%
  filter(year == year_min) %>%
  group_by(continent) %>%
  mutate(
    "lowerbound" = quantile(gdpPercap, 0.25) - IQR(gdpPercap)*1.5, 
    "upperbound" = quantile(gdpPercap, 0.75) + IQR(gdpPercap)*1.5
  ) %>%
  filter(gdpPercap < lowerbound | gdpPercap > upperbound)

outliers_1952

```

**Observations**:

- Identify the outlier countries from q2: Angola, Bahrain, Canada, Gabon, Kuwait, South Africa, Switzerland, United States, and Venezuela

__q4__ Create a plot similar to yours from q2 studying both `year_min` and `year_max`. Find a way to highlight the outliers from q3 on your plot. Compare the patterns between `year_min` and `year_max`.

*Hint*: We've learned a lot of different ways to show multiple variables; think about using different aesthetics or facets.

```{r q4-task}
## TASK: Create a visual of gdpPercap vs continent

gapminder %>%
  filter(year == year_min | year == year_max) %>%
  ggplot(aes(gdpPercap, as.factor(year), fill = as.factor(year))) +
    geom_boxplot() +
    geom_point(data = outliers_1952, color = "orange") +
    coord_trans(x = "log") +    
    facet_grid(continent ~ .) +
    labs(
      title = "GDP per capita by continent and year", 
      x = "GDP per capita", 
      y = "Year")


```

**Observations**:

- Mean GDP per capita increased for all continents from 1952 to 2007.
- In all cases except Asia, the boxplot's whiskers in 2007 extend higher than the GDP for the highest outlier in that continent in 1952.

# Your Own EDA
<!-- -------------------------------------------------- -->

Now it's your turn! We just went through guided EDA considering the GDP per capita at two time points. You can continue looking at outliers, consider different years, repeat the exercise with `lifeExp`, consider the relationship between variables, or something else entirely.

__q5__ Create *at least* three new figures below. With each figure, try to pose new questions about the data.

How do GDP and life expectancy compare?

```{r q5-task1}
## TASK: Your first graph
ggplot(gapminder, aes(gdpPercap, lifeExp)) +
  geom_point(aes(color = year)) +
  coord_trans(x = "log") +
  geom_smooth(color = "orange") +
  labs(
    title = "GDP per capita and life expectancy",
    x = "GDP per capita",
    y = "Life expectancy"
  )
```

**Observations**

- As we might expect, increasing GDP per capita tends to mean an increased life expectancy, 
except at the very top values of GDP per capita.
- I would hypothesize that these countries were likely oil-rich authoritarian states, with a high GDP benefiting only the wealthiest in the country, leaving a large portion of the population poor, and in correspondingly poor health. Or it's the US; this could really go either way.

```{r q5-task1 additional looking}
topGDP <-
  gapminder %>%
  filter(gdpPercap > 50000) %>%
  arrange(desc(gdpPercap))

topGDP
```

**Observations**
- All the GDP per capita values greater than $50,000 are Kuwait. (I'm feeling smug. All those political science classes are paying off right now.)
- At least the US didn't turn out to be in this set, I guess.


So... What's the trend in the US over time with respect to GDP and life expectancy?
```{r q5-task2}
## TASK: Your second graph
gapminder %>%
  filter(country == "United States") %>%
  ggplot(aes(year, gdpPercap)) +
    geom_point(aes(color = lifeExp)) +
    scale_color_continuous(name = "Life expectancy") +
    labs(
      title = "GDP per capita and life expectancy in the United States, 1952-2007",
      x = "Year",
      y = "GDP per capita"
    )
```

**Observations**

- Boring but good. 
- Now I'm curious about Kuwait though.


```{r q5-task3}
## TASK: Your third graph

gapminder %>%
  filter(country == "Kuwait") %>%
  ggplot(aes(year, gdpPercap)) +
    geom_point(aes(color = lifeExp)) +
    scale_color_continuous(name = "Life expectancy") +
    labs(
      title = "GDP per capita and life expectancy in Kuwait, 1952-2007",
      x = "Year",
      y = "GDP per capita"
    )
```


**Observations**

- Fascinating. Life expectancy increases pretty steadily over the time period, but meanwhile GDP falls dramatically from the early 1970s to the mid 1980s, then starts climbing again.
- I wonder what political changes were wrought in this time period to bring about this effect?





<!-- include-rubric -->
# Grading Rubric
<!-- -------------------------------------------------- -->

Unlike exercises, **challenges will be graded**. The following rubrics define how you will be graded, both on an individual and team basis.

## Individual
<!-- ------------------------- -->

| Category | Unsatisfactory | Satisfactory |
|----------|----------------|--------------|
| Effort | Some task __q__'s left unattempted | All task __q__'s attempted |
| Observed | Did not document observations | Documented observations based on analysis |
| Supported | Some observations not supported by analysis | All observations supported by analysis (table, graph, etc.) |
| Code Styled | Violations of the [style guide](https://style.tidyverse.org/) hinder readability | Code sufficiently close to the [style guide](https://style.tidyverse.org/) |

## Team
<!-- ------------------------- -->

| Category | Unsatisfactory | Satisfactory |
|----------|----------------|--------------|
| Documented | No team contributions to Wiki | Team contributed to Wiki |
| Referenced | No team references in Wiki | At least one reference in Wiki to member report(s) |
| Relevant | References unrelated to assertion, or difficult to find related analysis based on reference text | Reference text clearly points to relevant analysis |

## Due Date
<!-- ------------------------- -->

All the deliverables stated in the rubrics above are due on the day of the class discussion of that exercise. See the [Syllabus](https://docs.google.com/document/d/1jJTh2DH8nVJd2eyMMoyNGroReo0BKcJrz1eONi3rPSc/edit?usp=sharing) for more information.
