---
title: "Antibiotics"
author: "Angela Sharer"
date: 2020-07-26
output:
  github_document:
    toc: true
---

*Purpose*: To create an effective visualization, we need to keep our *purpose* firmly in mind. There are many different ways to visualize data, and the only way we can judge efficacy is with respect to our purpose.

In this challenge you'll visualize the same data in two different ways, aimed at two different purposes.

*Note*: Please complete your initial visual design **alone**. Work on both of your graphs alone, and save a version to your repo *before* coming together with your team. This way you can all bring a diversity of ideas to the table!

```{r setup}
library(tidyverse)
library(ggrepel)

```

*Background*: The data[1] we study in this challenge report the [*minimum inhibitory concentration*](https://en.wikipedia.org/wiki/Minimum_inhibitory_concentration) (MIC) of three drugs for different bacteria. The smaller the MIC for a given drug and bacteria pair, the more practical the drug is for treating that particular bacteria. An MIC value of *at most* 0.1 is considered necessary for treating human patients.

These data report MIC values for three antibiotics---penicillin, streptomycin, and neomycin---on 16 bacteria. Bacteria are categorized into a genus based on a number of features, including their resistance to antibiotics.

```{r load}
## NOTE: If you extracted all challenges to the same location,
## you shouldn't have to change this filename
filename <- "./data/antibiotics.csv"

## Load the data
df_antibiotics <- read_csv(filename)
df_antibiotics %>% knitr::kable()
```


**What's the deal with this "gram" column?**

According to Wikipedia, "Gram staining differentiates bacteria by the chemical and physical properties of their cell walls. Gram-positive cells have a thick layer of peptidoglycan in the cell wall... Gram-negative cells have a thinner peptidoglycan layer." This can make them respond differently to e.g. antibiotics. https://en.wikipedia.org/wiki/Gram_stain


# Visualization
<!-- -------------------------------------------------- -->

## Purpose: Compare Effectiveness
<!-- ------------------------- -->

```{r df modification for visualization}
df_antibiotics2 <-
  df_antibiotics %>%
  group_by(bacteria) %>%
  mutate(min_MIC = min(penicillin, streptomycin, neomycin)) %>%
  ungroup() %>%
  pivot_longer(
    names_to = "drug",
    values_to = "MIC",
    cols = c(penicillin, streptomycin, neomycin)
  ) %>%
  mutate(
    effective = MIC <= 0.1
    ) %>%
  separate(
    col = bacteria, 
    into = c("genus","species1", "species2"), 
    sep = " ",
    remove = FALSE
    ) %>%
  unite(col = species, species1, species2, sep = " ", na.rm = TRUE)




df_antibiotics2 
```

__q1__ Create a visualization of `df_antibiotics` that helps you to compare the effectiveness of the three antibiotics across all the bacteria reported. Can you make any broad statements about antibiotic effectiveness?

```{r q1-task boxplot antibiotic vs MIC}
## TASK: Create your visualization
df_antibiotics2 %>%
  ggplot(aes(drug, MIC)) +
  geom_boxplot() +
  coord_trans(y = "log") +
  geom_hline(yintercept = 0.1, linetype = "dashed", color = "blue") +
  labs(
    title = "MIC of 3 antibiotics against various bacteria",
    x = "Antibiotic",
    y = "Minimum inhibitory concentration (MIC)"
  )
```

**Observations**:

- The effectiveness of penicillin varies more than the other two antibiotics.
- Neomycin seems to be more effective on average than either of the other drugs (that is, it has the lowest average MIC value against the bacteria in this dataset). 


```{r q1-task faceted col plot in the effective range}
df_antibiotics2 %>%
  ggplot(aes(MIC, drug)) +
  geom_col(aes(fill = effective)) +
  scale_fill_discrete(name = "Effective for humans?") +
  coord_cartesian(xlim = c(0, 0.1)) +
  facet_wrap(~ bacteria) +
  labs(
    title = "MIC of 3 antibiotics against each bacteria",
    x = "Minimum inhibitory concentration (MIC)",
    y = "Antibiotic"
  )
## I've got a ton of clipping and overlapping text happening here. Not sure how to fix.
```

**Observations**:

- The antibiotics investigated here cannot effectively combat all of these bacteria in humans. In fact, these antibiotics are only effective against 12 of the 16 bacteria here. 


```{r q1-task jitter plot MIC vs bacteria}
df_antibiotics2 %>%
  ggplot(aes(MIC, bacteria, color = drug)) +
  geom_jitter(width = 0, height = 0.15) +
  coord_trans(x = "log") + 
  scale_color_discrete(name = "Antibiotic") +
  geom_vline(xintercept = 0.1, linetype = "dashed", color = "blue") +
  labs(
    title = "MIC of 3 antibiotics against each bacteria",
    x = "Minimum inhibitory concentration (MIC)",
    y = "Bacteria"
  ) 
```

**Observations**:

- Each of these antibiotics varies substantially in its effectiveness, depending on the bacteria in question. 



## Purpose: Categorize Bacteria
<!-- ------------------------- -->

The *genus* of a living organism is a human categorization, based on various characteristics of the organism. Since these categories are based on numerous factors, we will tend to see clusters if we visualize data according to relevant variables. We can use these visuals to categorize observations, and to question whether given categories are reasonable!

__q2__ Create a visualization of `df_antibiotics` that helps you to categorize bacteria according to the variables in the data. Document your observations on how how clusters of bacteria in the variables do---or don't---align with their *genus* classification.

```{r q2-task jitter plot genus vs MIC}
## TASK: Create your visualization
df_antibiotics2 %>%
  ggplot(aes(MIC, genus, color = drug)) +
  geom_jitter(width = 0, height = 0.15) +
  coord_trans(x = "log") + 
  scale_color_discrete(name = "Antibiotic") +
  geom_vline(xintercept = 0.1, linetype = "dashed", color = "blue") +
  labs(
    title = "MIC of 3 antibiotics against each genus of bacteria",
    x = "Minimum inhibitory concentration (MIC)",
    y = "Genus of bacteria"
  ) 
```
**Observations**:

- In many cases, there was only one species of bacteria representing the entire genus. 
- I don't think my plot, above, is very useful.


```{r q2-task jitter plot antiobiotic vs. MIC, genus}
df_antibiotics2 %>%
  ggplot(aes(drug, MIC)) +
  geom_jitter(aes(color = genus), height = 0, width = 0.15) +
  coord_trans(y = "log") +
  geom_hline(yintercept = 0.1, linetype = "dashed", color = "blue") +
  labs(
    title = "MIC of 3 antibiotics against various bacteria",
    x = "Antibiotic",
    y = "Minimum inhibitory concentration (MIC)"
  )
```

**Observations**:

- This also isn't very helpful.


```{r q2-task jiitter plot, antibiotic vs. MIC, gram}
df_antibiotics2 %>%
  ggplot(aes(drug, MIC)) +
  geom_jitter(aes(color = gram), height = 0, width = 0.15) +
  coord_trans(y = "log") +
  geom_hline(yintercept = 0.1, linetype = "dashed", color = "blue") +
  labs(
    title = "MIC of 3 antibiotics against various bacteria",
    x = "Antibiotic",
    y = "Minimum inhibitory concentration (MIC)"
  )

```

**Observations**:

- These antibiotics are not effective against most of the gram-negative bacteria in this dataset.
- However, these antibiotics are effective against most of the gram-positive bacteria in this dataset. 


```{r q2-task boxplot antibiotic vs MIC, gram}
df_antibiotics2 %>%
  ggplot(aes(drug, MIC)) +
  geom_boxplot() +
  coord_trans(y = "log") +
  geom_hline(yintercept = 0.1, linetype = "dashed", color = "blue") +
  facet_grid(. ~ gram) +
  labs(
    title = "MIC of 3 antibiotics against various bacteria",
    x = "Antibiotic",
    y = "Minimum inhibitory concentration (MIC)"
  )
```

**Observations**:

- The gram type (positive or negative) matters: Penicillin is much more effective against gram-positive bacteria, and neomycin and streptomycin have potential to be so as well (though their effectiveness against the gram-positive bacteria in this dataset varies more than pencillin's does).
- These antibiotics work better on gram-positive bacteria, it seems; and only neomycin is effective for use in humans against a (particular) gram-negative bacteria. 

# References
<!-- -------------------------------------------------- -->

[1] Neomycin in skin infections: A new topical antibiotic with wide antibacterial range and rarely sensitizing. Scope. 1951;3(5):4-7.

[2] Wainer and Lysen, "That's Funny..." /American Scientist/ (2009) [link](https://www.americanscientist.org/article/thats-funny)


<!-- include-rubric -->
# Grading Rubric
<!-- -------------------------------------------------- -->

Unlike exercises, **challenges will be graded**. The following rubrics define how you will be graded, both on an individual and team basis.

## Individual
<!-- ------------------------- -->

| Category | Unsatisfactory | Satisfactory |
|----------|----------------|--------------|
| Effort | Some task __q__'s left unattempted | All task __q__'s attempted |
| Observed | Did not document observations | Documented observations based on analysis |
| Supported | Some observations not supported by analysis | All observations supported by analysis (table, graph, etc.) |
| Code Styled | Violations of the [style guide](https://style.tidyverse.org/) hinder readability | Code sufficiently close to the [style guide](https://style.tidyverse.org/) |

## Team
<!-- ------------------------- -->

| Category | Unsatisfactory | Satisfactory |
|----------|----------------|--------------|
| Documented | No team contributions to Wiki | Team contributed to Wiki |
| Referenced | No team references in Wiki | At least one reference in Wiki to member report(s) |
| Relevant | References unrelated to assertion, or difficult to find related analysis based on reference text | Reference text clearly points to relevant analysis |

## Due Date
<!-- ------------------------- -->

All the deliverables stated in the rubrics above are due on the day of the class discussion of that exercise. See the [Syllabus](https://docs.google.com/document/d/1jJTh2DH8nVJd2eyMMoyNGroReo0BKcJrz1eONi3rPSc/edit?usp=sharing) for more information.
