---
title: "Team Zeta Project 1"
author: "Angela Sharer"
date: 2020-08-03
output:
  github_document:
    toc: true
---

# Food Consumption and CO2 Emissions
Team Zeta Project 1

## Setup

```{r setup}
library(tidyverse)

food_consumption <- readr::read_csv('https://raw.githubusercontent.com/rfordatascience/tidytuesday/master/data/2020/2020-02-18/food_consumption.csv')

gapminder_pop_total <- read_csv("./data/population_total.csv")

gapminder_geo <- read_csv("./data/countries_gapminder.csv")

gapminder_emissions <- read_csv("./data/co2_emissions_tonnes_per_person.csv")

```


```{r examining data sets}
food_consumption

gapminder_pop_total

gapminder_geo

gapminder_emissions
```

## Data Wrangling

```{r gapminder wrangling}
gapminder_pop_2014 <-
  gapminder_pop_total %>%
  select("country", "population" = "2014")

gapminder_pop_2014

gapminder_geo_select <-
  gapminder_geo %>%
  select(
    "country" = "name", 
    "region" = "World bank region", 
    "income_grp" = "World bank, 4 income groups 2017"
  )

gapminder_geo_select

gapminder_emissions_2014 <-
  gapminder_emissions %>%
  select("country", "total_co2_tonnes" = "2014") %>%
  mutate(total_co2_emissions = total_co2_tonnes * 1000) %>%
  select("country", "total_co2_emissions")

gapminder_emissions_2014

```

```{r food_consumption wrangling - fixing countries}
## First, I'm fixing the USA
food_consumption_US_only <-
  food_consumption %>%
  filter(country == "USA") %>%
  select(-country) %>%
  mutate(country = "United States") %>%
  select(country, food_category, consumption, co2_emmission)

## Next, I'm fixing Hong Kong
food_consumption_HK_only <-
  food_consumption %>%
  filter(country == "Hong Kong SAR. China") %>%
  select(-country) %>%
  mutate(country = "Hong Kong, China") %>%
  select(country, food_category, consumption, co2_emmission)

## Next, I'm fixing Taiwan
food_consumption_TW_only <-
  food_consumption %>%
  filter(country == "Taiwan. ROC") %>%
  select(-country) %>%
  mutate(country = "Taiwan") %>%
  select(country, food_category, consumption, co2_emmission)

## Macedonia (Macedonia, FYR in gapminder)
food_consumption_Macedonia_only <-
  food_consumption %>%
  filter(country == "Macedonia") %>%
  select(-country) %>%
  mutate(country = "Macedonia, FYR") %>%
  select(country, food_category, consumption, co2_emmission)

## Congo, Rep.
food_consumption_CongoRep_only <-
  food_consumption %>%
  filter(country == "Congo") %>%
  select(-country) %>%
  mutate(country = "Congo, Rep.") %>%
  select(country, food_category, consumption, co2_emmission)


## Now, I'm binding them all together!
food_consumption_mod <-
  food_consumption %>%
  filter(
    country != "USA" &
    country != "Hong Kong SAR. China" &
    country != "Taiwan. ROC" &
    country != "Macedonia" &
    country != "Congo"
  ) %>%
  bind_rows(
    food_consumption_US_only,
    food_consumption_HK_only,
    food_consumption_TW_only,
    food_consumption_Macedonia_only,
    food_consumption_CongoRep_only
    )
```

```{r food_consumption wrangling - joining with gapminder}
df_food_pop <-
  food_consumption_mod %>%
  left_join(gapminder_pop_2014, by = "country") %>%
  left_join(gapminder_geo_select, by = "country") %>%
  left_join(gapminder_emissions_2014, by = "country") %>%
  group_by(country) %>%
  mutate(
    co2_food_country = sum(co2_emmission), 
    food_consumption_country = sum(consumption)
  ) %>%
  ungroup() %>%
  select(
    region,
    country,
    population,
    income_grp,
    food_category,
    consumption,
    food_consumption_country,
    "co2_emission_food" = co2_emmission,
    co2_food_country,
    total_co2_emissions
  )

df_food_pop

#View(df_food_pop[is.na(df_food_pop$region),])
## Gapminder is missing these countries, which are present in food_consumption:
## Bermuda, French Polynesia, New Caledonia, Slovakia
```

## EDA

```{r scatterplot of consumption and co2 by country and region}
df_food_pop %>%
  ggplot() +
  geom_point(aes(co2_food_country, food_consumption_country, color = region)) +
  scale_color_discrete(name = "Region") +
  labs(
    title = "Food consumption and CO2 emissions per capita",
    x = "CO2 emissions per capita (kg CO2/person/year)",
    y = "Food consumption per capita (kg/person/year)"
  )
```


```{r scatterplot of consumption and co2 by country and income group}
df_food_pop %>%
  ggplot() +
  geom_point(aes(co2_food_country, food_consumption_country, color = income_grp)) +
  scale_color_discrete(name = "Income level") +
  labs(
    title = "Food consumption and CO2 emissions per capita",
    x = "CO2 emissions per capita (kg CO2/person/year)",
    y = "Food consumption per capita (kg/person/year)"
  )
```


```{r boxplot of consumption by food category}
df_food_pop %>%
  ggplot() +
  geom_boxplot(aes(consumption, fct_reorder(food_category, consumption))) +
  labs(
    title = "Consumption per capita by food category",
    x = "Food consumption per capita (kg/person/year)",
    y = "Food Category"
  )
```
```{r facetted boxplot of consumption by food category and region}
df_food_pop %>%
  filter(!is.na(region)) %>%
  ggplot() +
  geom_boxplot(aes(consumption, fct_reorder(food_category, consumption))) +
  facet_wrap(~ fct_reorder(region, desc(consumption)), ncol = 4) +
  labs(
    title = "Consumption per capita by food category and region",
    x = "Food consumption per capita (kg/person/year)",
    y = "Food Category"
  )
```



```{r facetted boxplot of consumption by food category and income group}
df_food_pop %>%
  filter(!is.na(income_grp)) %>%
  ggplot() +
  geom_boxplot(aes(consumption, fct_reorder(food_category, consumption))) +
  facet_wrap(~ fct_reorder(income_grp, desc(consumption)), ncol = 4) +
  labs(
    title = "Consumption per capita by food category and income group",
    x = "Food consumption per capita (kg/person/year)",
    y = "Food Category"
  )
```

```{r food emissions vs. total emissions}
df_food_pop %>%
  ggplot() +
  geom_point(aes(co2_food_country, total_co2_emissions, color = region), na.rm = TRUE) +
  scale_color_discrete(name = "Region") +
  #coord_trans(x = "log", y = "log") +
  labs(
    title = "CO2 emissions per capita from food vs. overall",
    x = "CO2 emissions from food per capita (kg CO2/person/year)",
    y = "Total CO2 emissions per capita (kg CO2/person/year)"
  )

```

```{r rice vs. wheat}

df_rice_wheat <-
  df_food_pop %>%
  filter(
    food_category == "Wheat and Wheat Products" | 
    food_category == "Rice"
  ) %>%
  select(-co2_emission_food) %>%
  pivot_wider(
    names_from = food_category,
    values_from = consumption
  ) %>%
  select(
    region,
    country,
    income_grp,
    food_consumption_country,
    "wheat" = "Wheat and Wheat Products",
    "rice" = "Rice"
  )

df_rice_wheat %>%
  ggplot() +
  geom_point(
    aes(
      wheat, 
      rice, 
      color = fct_reorder(income_grp, desc(food_consumption_country))
    )
  ) +
  scale_color_discrete(name = "Income level") +
  labs(
    title = "Wheat and Rice consumption per capita",
    x = "Wheat and wheat product consumption per capita",
    y = "Rice consumption per capita"
  )

```

```{r dairy; Finland}

df_food_pop %>%
  filter(food_category == "Milk - inc. cheese") %>%
  arrange(desc(consumption))

df_food_pop %>%
  filter(country == "Finland") %>%
  ggplot() +
  geom_col(aes(fct_reorder(food_category, desc(consumption)), consumption)) +
  scale_x_discrete(
    labels = function(food_category) str_wrap(food_category, width = 10)
  ) +
  labs(
    title = "Finland's food consumption per capita, select categories",
    x = "Food category",
    y = "Consumption per person per year"
  )

```

This almost seems like a mistake. I have asked my friend in Finland whether Finns consume a lot of milk and/or cheese, and will report back when she responds.

The internet claims Finns drink a lot of milk. Interesting.




*Other ideas for graphs:*

- Milk vs. GDP
- Rice vs. GDP
- Worst emitters breakdown
- Lowest emitters breakdown




# Sources

We downloaded the data on food consumption and CO2 emissions from Tidy Tuesday on Github, but that data came from another source, who originally got it from the Food and Agriculture Organization of the United Nations.

- [Food consumption data, tidy format](https://github.com/rfordatascience/tidytuesday/blob/master/data/2020/2020-02-18/readme.md)
- [Food consumption data, source](https://www.nu3.de/blogs/nutrition/food-carbon-footprint-index-2018)
- [Food consumption data, original source](http://www.fao.org/faostat/en/#data)

We merged in some additional data on population, income, etc. from Gapminder.

- [Gapminder](https://www.gapminder.org/data/)